Differences from our PPoPP Conference Paper \cite{KBP16}

### 2018-06-22 latex/toms/experiments.tex
sc papers in toms format

\section{Experiments}
\label{sec:experiment}
\newcommand{\NLS}{LUC }

In this section, we describe our implementation of \ParNMF\ and evaluate its performance.
We identify a few synthetic and real world data sets to experiment with \ParNMF\ with dimensions that span from hundreds to millions. 
We compare the performance and exploring scaling behavior of different NMF algorithms -- \MU, \HALS, and ANLS/BPP (\BPP), implemented using the parallel \ParNMF\ framework.  
The code and the  datasets used for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.

\subsection{Experimental Setup}

\subsubsection{Data Sets}\label{sec:datasets}

We used sparse and dense matrices that are either synthetically generated or from real world applications. We explain the data sets in this section.

\begin{itemize}
\item Dense Synthetic Matrix: We generate a low rank matrix as the product of two uniform random matrices of size 207,360 $\times$ 100 and 100 $\times$ 138,240. 
The dimensions of this matrix are chosen to be evenly divisible for a particular set of processor grids.  
\item Sparse Synthetic Matrix: We generate a random sparse Erd\H{o}s-R\'{e}nyi matrix of the size 207,360 $\times$ 138,240 with density of 0.001.  That is, every entry is nonzero with probability 0.001.
\item Dense Real World Matrix ({\em Video}):  NMF is used on video data for background subtraction in order to detect moving objects. The low rank matrix $\hat{\AA} = \WW \HH$ represents background and the error matrix $\AA - \hat{\AA}$ represents moving objects.  %Detecting moving objects has many real-world applications such as traffic estimation \cite{Fujimoto2014} and security monitoring \cite{BSJJZ2015}.
In the case of detecting moving objects in streaming videos, the last several minutes of video is taken from the live video camera to construct the non-negative matrix. 
An algorithm to incrementally adjust the NMF based on the streaming video is presented in \cite{kim2013nonnegative}. 
To simulate this scenario, we collected a video in a busy intersection of the Georgia Tech campus at 20 frames per second. 
From this video, we took video for approximately 12 minutes and  then reshaped the matrix such that every RGB frame is a column of our matrix, so that the matrix is dense with size 1,013,400 $\times$ 13,824. 
\item Sparse Real World Matrix ({\em Webbase}): This data set is a directed sparse graph whose nodes correspond to webpages (URLs) and edges correspond to hyperlinks from one webpage to another.
The NMF output of this directed graph helps us understand clusters in graphs.
We consider two versions of the data set: {\em webbase-1M} and {\em webbase-2001}.
The dataset webbase-1M contains about 1 million nodes (1,000,005) and 3.1 million edges (3,105,536), and was first reported by Williams et al. \cite{Williams2009}.  
The version webbase-2001 has about 118 million nodes (118,142,155) and over 1 billion edges (1,019,903,190); it was first reported by Boldi and Vigna  \cite{Boldi2004}.  
Both data sets are available in the University of Florida Sparse Matrix Collection \cite{DH11} and the latter {\em webbase-2001} being the largest among the entire collection.
\item Text data ({\em Stack Exchange}): 
Stack Exchange is a network of question-and-answer websites on topics in varied fields, each site covering a specific topic, where questions, answers, and users are subject to a reputation award process. There are many Stack Exchange forums, such as {\em ask ubuntu, mathematics, latex}. 
We downloaded the latest anonymized dump of all 
user-contributed content on the Stack Exchange network from \cite{stkxchngdataset}. 
%Each site is formatted as a separate archive consisting of XML files zipped via 7-zip using bzip2 compression. 
%Each site archive includes Posts, Users, Votes, Comments, PostHistory and PostLinks. 
%For complete schema information, see the included readme.txt. 
We used only the questions from the most popular site called Stackoverflow and did not include the answers and comments. 
We removed the standard 571 English stop words (such as {\em are, am, be, above, below}) and then used snowball stemming available through the Natural Language Toolkit (NLTK) package \cite{LB2002}.
After this initial pre-processing, we deleted HTML tags (such as {\em lt, gt, em}) from the posts. 
The resulting bag-of-words matrix has a vocabulary of size 627,047 over 11,708,841 documents with 365,168,945 non-zero entries. 
In this data, the vocabulary is larger than the typical set of English words because it includes variables, constants, and other programming constructs of various programming languages from the user questions.
%We ran NMF on this dataset for finding 50 topics whose sample topic words are presented at the end of this section. 
\end{itemize}

The size of all the real world data sets were adjusted to the nearest size for uniformly distributing the matrix.

\subsubsection{Implementation Platform}

We conducted our experiments on ``Rhea'' at the Oak Ridge Leadership Computing Facility (OLCF).
Rhea is a commodity-type Linux cluster with a total of 512 nodes and a 4X FDR Infiniband interconnect.
Each node contains dual-socket 8-core Intel Sandy Bridge-EP processors and 128 GB of memory.
Each socket has a shared 20MB L3 cache, and each core has a private 256K L2 cache.

%\grey{Do we want to start using the acronym for our software and discuss it here? We will park this for camera ready}

Our objective of the implementation is using open source software as much as possible 
to promote reproducibility and reuse of our code.
The entire C++ code is developed using the matrix library Armadillo \cite{sanderson2010}. 
In Armadillo, the elements of the dense matrix are stored in column major order and the sparse matrices in Compressed Sparse Column (CSC) format.
For dense BLAS and LAPACK operations, we linked Armadillo with Intel MKL -- the default LAPACK/BLAS library in RHEA. It is also easy to link Armadillo with OpenBLAS \cite{xianyi2015}. 
We use Armadillo's own implementation of sparse matrix-dense matrix multiplication, the default GNU C++ Compiler (g++ (GCC) 4.8.2) and MPI library (Open MPI 1.8.4)  on RHEA.  We chose the commodity cluster with open source software so that the numbers presented here are representative of common use.

%\grey{Need to update this last phrase with compilation setup on Rhea}.

\subsubsection{Algorithms}

In our experiments, we considered the following algorithms: 
\begin{itemize}
	\item \MU: \ParNMF\ (Algorithm \ref{alg:2D}) with MU (Equation \eqref{eqn:muupdate})
	\item \HALS: \ParNMF\ (Algorithm \ref{alg:2D}) with HALS (Equation \eqref{eqn:halsupdate})
	\item \BPP: \ParNMF\ (Algorithm \ref{alg:2D}) with BPP (Section \ref{sec:BPP})
	\item \Naive: \NaiveAlg\ (Algorithm \ref{alg:naive}, Section \ref{sec:naive})
\end{itemize}

Our implementation of \Naive\ (Algorithm \ref{alg:naive}) uses BPP but can be easily to extended to \MU, \HALS, and other NMF algorithms. 
%A detailed comparison of \NaiveAlg\ with \ParNMF\ is made in our earlier work \cite{KBP16}. 
%We include some benchmark results from \Naive\ to reiterate the point that communication efficiency is key to obtaining reasonable performance, but we also omit other \Naive\ results in order to focus attention on comparisons among other algorithms.

For the algorithms based on \ParNMF, we use the processor grid that is closest to the theoretical optimum (see Section \ref{sec:alg:comm}) in order to minimize communication costs.
See Section \ref{sec:procgrid} for an empirical evaluation of varying processor grids for a particular algorithm and data set.

%We choose these three algorithms to confirm the following conclusions from the analysis of Section \ref{sec:parNMF}: the performance of a naive parallelization of \NaiveAlg\ (Algorithm \ref{alg:naive}) will be severely limited by communication overheads, and the correct choice of processor grid within Algorithm \ref{alg:2D} is necessary to optimize performance.
%To demonstrate the latter conclusion, we choose the two extreme choices of processor grids and test some data sets where a 1D processor grid is optimal (e.g., the Video matrix) and some where a squarish 2D grid is optimal (e.g., the Webbase matrix).

To ensure fair comparison among algorithms, the same random seed is used across different methods appropriately. 
That is, the initial random matrix $\HH$ is generated with the same random seed when testing with 
different algorithms (note that $\WW$ need not be initialized). 
In our experiments, we use number of iterations as the stopping criteria for all the algorithms.

%\grey{Need to update the following paragraph for our running times, maybe consider moving it elsewhere...}
While we would like to compare against other high-performance NMF algorithms in the literature, the only other distributed-memory implementations of which we're aware are implemented using Hadoop and are designed only for sparse matrices \cite{liao2014cloudnmf},
\cite{liu2010distributed}, \cite{gemulla2011large}, \cite{Yin2014} and \cite{Faloutsos2014}.
We stress that Hadoop is not designed for high performance computing of iterative numerical 
algorithms, requiring disk I/O between steps, so a run time comparison between a Hadoop 
implementation and a C++/MPI implementation is not a fair comparison of parallel algorithms.
A qualitative example of differences in run time is that a Hadoop implementation of the MU algorithm on 
a large sparse matrix of size $2^{17} \times 2^{16}$ with $2 \times {10^8}$ nonzeros (with k=8) 
takes on the order of 50 minutes per iteration \cite{liu2010distributed}, while our MU implementation 
takes 0.065 seconds per iteration for the synthetic data set (which is an order of magnitude larger in 
terms of rows, columns, and nonzeros) running on only 16 nodes. 
%Based on our survey, we are the fastest distributed NMF algorithm in the literature.

% input file with plotting macros
\input{plots.tex}

\subsection{Relative Error over Time} \label{sec:convergence}

There are various metrics to compare the quality of the 
NMF algorithms \cite{kim2013nonnegative}. The most common among these metrics are (a) relative error and (b) projected 
gradient. The former represents the closeness of the low rank approximation $\hat{\AA}\approx\WW\HH$, which is generally the optimization objective. 
The latter 
represent the quality of the produced low rank factors and the stationarity of the final solution. These 
metrics are also used as the stopping criterion for terminating the iteration of the NMF algorithm as in 
line \ref{algo:nmfloop} of Algorithm \ref{alg:aunmf}. Typically a combination of the number of iterations 
along with improvement of these metrics until a tolerance is met is used as stopping criterion. In this paper, we use 
relative error for the comparison as it is monotonically decreasing, as opposed to projected gradient of the 
low rank factors, which shows oscillations over iterations. The relative error can be formally defined as 
$\|\AA-\WW\HH\|_F/\|\AA\|_F$.

In Figure \ref{fig:convergence}, we measure the relative error at the end of every iteration (i.e., after the updates of both $\WW$ and $\HH$) for all three algorithms \MU, \HALS, and 
\BPP, and we plot the relative error over time (each mark represents an iteration).
We consider three real world datasets, \emph{video}, \emph{stack exchange} and \emph{webbase-1M}, and set $k=50$. 
We used only the number of iterations as stopping criterion and,  
just for this section, ran all the algorithms for 30 iterations. 
We note that the convergence behavior and computed factors can vary over different initializations; we used the same initial values across all three algorithms in these experiments.
Also, we observed that for these data sets, the convergence behavior was not sensitive to initialization (the final residual errors varied by less than 1\% in our experiments).
NMF solutions are guaranteed to be unique in certain cases, with mild assumptions on the input data \cite{BLRG14,HFS2016}, but we do not check those assumptions for these datasets.

To begin with, we explain the observations on the dense \emph{video} dataset presented in Figure \ref{fig:denserwerr}. 
The relative error of \MU\ is highest at 0.1812 after 30 iterations and \HALS's is the least with 0.1273; \BPP's relative error
is  0.1716 after 30 iterations. 
%For this data, although the per-iteration times vary across algorithms and \BPP\ has the longest per-iteration time, it converged more quickly than \HALS\ or \MU\ because it solves subproblems exactly.
%From the figure, we can observe that \BPP\  error didn't change after 29 iterations where as \HALS\ and \MU\ was still improving marginally at the 4th decimal even after 30 iterations.

We can observe that the relative error of \emph{stack exchange} from Figure \ref{fig:stackexchangeerr} is better 
than \emph{webbase-1M} from Figure \ref{fig:sparserwerr} over all three algorithms. 
In the case of the \emph{stack exchange} dataset, the relative errors after 30 iterations follow the pattern \MU\ $>$ \HALS\ $>$ \BPP, with values 0.8509, 0.8395, and 0.8377 respectively. 
%Unlike the \emph{video} dataset, both \MU\ and \HALS\ stopped improving after 23 iterations, 
%where as \BPP\ was still improving in the 4th decimal even though its error was better than the others. 
However, the difference in relative error for the \emph{webbase-1M} dataset is negligible, though the relative ordering of \MU\ $>$ \HALS\ $>$ \BPP\ is consistent, with values of 0.99927 for \MU\, 0.99920 for \HALS\ and 0.99919 for \BPP.

In general, for these datasets \BPP\ identifies better approximations and converges faster than \MU\ and \HALS\ despite the extra per-iteration time, which is consistent with the 
literature \cite{kim2013nonnegative,kim2011fast}. 
However, for the sparse datasets, the differences in relative error are small across the NMF algorithms.

%The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}.

%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont,singlelinecheck=off,justification=raggedright}

\begin{figure}

% set which run to plot
\renewcommand{\run}{3}

\begin{subfigure}{0.3 \columnwidth}
\ylabeltrue
\begin{tikzpicture}
\renewcommand{\datafile}{data/denserwerr-time.dat}
\renewcommand{\run}{1}
\relerrplot
\renewcommand{\run}{3}
\end{tikzpicture}
\ylabelfalse
\subcaption{Video}
\label{fig:denserwerr}
\end{subfigure}
~
\begin{subfigure}{0.3 \columnwidth}
\begin{tikzpicture}
\renewcommand{\datafile}{data/stkx-5runs-err.dat}
\legendtrue
\relerrplot
\legendfalse
\end{tikzpicture}
\subcaption{Stack Exchange}
\label{fig:stackexchangeerr}
\end{subfigure}
~
\begin{subfigure}{0.3 \columnwidth}
\begin{tikzpicture}
\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
\relerrplot
\end{tikzpicture}
\subcaption{Webbase}
\label{fig:sparserwerr}
\end{subfigure}

\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
\label{fig:convergence}
\end{figure}

\subsection{Time Per Iteration}

In this section we focus on per-iteration time of all the algorithms.
We report four types of experiments, varying the number of processors (Section \ref{sec:scaling}), the rank of the approximation (Section \ref{sec:ksweep}), the shape of the processor grid (Section \ref{sec:procgrid}), and scaling up the dataset size.
For each experiment we report a time breakdown in terms of the overall computation and communication steps (described in Section \ref{sec:perf-breakdown}) shared by all algorithms.

\subsubsection{Time Breakdown}
\label{sec:perf-breakdown}

To differentiate the computation and communication costs among the algorithms, we present the time breakdown among the various tasks within the algorithms for all performance experiments.
For Algorithm \ref{alg:2D}, there are three local computation tasks and three communication tasks to compute each of the factor matrices:
\begin{itemize}
	\item \textbf{MM}, computing a matrix multiplication with the local data matrix and one of the factor matrices;
	\item \textbf{\LUC}, local updates either using \BPP\ or applying the remaining work of the \MU\ or \HALS\ updates (i.e., the total time for both $UpdateW$ and $UpdateH$ functions); 
	\item \textbf{Gram}, computing the local contribution to the Gram matrix;
	\item \textbf{All-Gather}, to compute the global matrix multiplication;
	\item \textbf{Reduce-Scatter}, to compute the global matrix multiplication;
	\item \textbf{All-Reduce}, to compute the global Gram matrix.
\end{itemize}
In our results, we do not distinguish the costs of these tasks for $\WW$ and $\HH$ separately; we report their sum, though we note that we do not always expect balance between the two contributions for each task.
Algorithm \ref{alg:naive} performs all of these tasks except Reduce-Scatter and All-Reduce; all of its communication is in All-Gather.

\subsubsection{Scaling \texorpdfstring{$p$}{p}: Strong Scaling}
\label{sec:scaling}

Figure \ref{fig:scaling} presents a strong scaling experiment with four data sets: \emph{sparse synthetic}, \emph{dense synthetic}, \emph{webbase-1M}, and \emph{video}.
In this experiment, for each data set and algorithm, we use low rank $k=50$ and vary the number of processors (with fixed problem size).
We use $\{1,6,24,54,96\}$ nodes; since each node has 16 cores, this corresponds to $\{16,96,384,864,1536\}$ cores.
We report average per-iteration times.

We highlight three main observations from these experiments:
\begin{enumerate}
	\item \label{obs:1} \Naive\ is slower than all other algorithms for large $p$;
	\item \label{obs:2} \MU, \HALS, and \BPP\ (algorithms based on \ParNMF) scale up to over 1000 processors;
	\item \label{obs:3} the relative per-iteration cost of \LUC\ decreases as $p$ increases (for all algorithms), and therefore the extra per-iteration cost of \BPP\ (compared with \MU\ and \HALS) becomes negligible.
\end{enumerate}

\paragraph{Observation \ref{obs:1}} 
%We report \Naive\ performance only for the synthetic data sets (Figures \ref{fig:sparsesynstrongscale} and \ref{fig:densesynscaling}); the results for the real-world data sets are similar.
For the Sparse Synthetic data set, \Naive\ is $4.2\times$ slower than the fastest algorithm (\BPP) on 1536 processors; for the Dense Synthetic data set, \Naive\ is $1.6\times$ slower than the fastest algorithm (\MU) at that scale.
The slowdown increases to $7.7\times$ and $3.6\times$ for the sparse and dense real-world datasets, respectively.
Nearly all of this slowdown is due to the communication costs of \Naive.
Theoretical and practical evidence supporting the first observation is also reported in our previous paper \cite{KBP16}.
However, we also note that \Naive\ is the fastest algorithm for the smallest $p$ for each problem, which is largely due to reduced MM time.
Each algorithm performs exactly the same number of flops per MM; the efficiency of \Naive\ for small $p$ is due to cache effects.
For example, for the Dense Synthetic problem on 96 processors, the output matrix of \Naive's MM fits in L2 cache, but the output matrix of \ParNMF's MM does not; these effects disappear as $p$ increases.

>>>>

% !TEX root = paper.tex

\section{Performance Results} \label{sec:experiments}

\input{plots}

\subsection{Datasets}

\subsubsection{Hyperspectral Images (HSI)}
For comparison with previous work \cite{LK+17b}, we consider the same 3D hyperspectral imaging dataset called ``Souto\_wood\_pile'' \cite{FAN16}. 
NNCP is often used on HSI data sets for classification and blind source separation of materials with differing spectral signatures.
The hyperspectral datacube has dimensions $1024 \times 1344 \times 33$ and represents a set of 33 grayscale images of size 1344 $\times$ 1024 pixels sampled at wavelengths 400, 410, $\dots$, 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$. 
We also consider the Nogueir\'{o} scene dataset, which is a sequence of 9 time-lapse HSI images of the same scene acquired at about 1-hour intervals.
In each scene, hyperspectral images were acquired at about 1-hour intervals. 
Each Nogueir\'{o} scene HSI image has the same properties as the Souto\_wood\_pile data set, so the corresponding tensor has dimensions $1024 \times 1344 \times 33 \times 9$.

\subsubsection{Dynamic Functional Connectivity (dFC)}
We also consider dynamic functional connectivity datasets that are generated from fMRI images of human brains.
Given a 4D fMRI data set of voxel measurements across multiple timesteps, voxels containing brain data are partitioned into a set of regions of interest (specified using domain-specific knowledge), and a single time-series signal is aggregated for each region of interest.
Then, an instantaneous correlation is computed for each time point and pair of regions, and this process is repeated for a number of subjects.
Computing a CP decomposition of this data helps to discover patterns of brain connectivity among different regions and also differentiate among individuals.
For our representative dFC data set, we consider 246 brain regions, which yields 30{,}012 unique pairs of regions, 1200 times steps, and 500 subjects, or a tensor of dimension $30{,}012\times 1200 \times 500$ \cite{VEU+12,THBGW17}.

\subsubsection{Synthetic}
Our synthetic data sets are constructed from a CP model with an exact low rank with no added noise.
In this case we can confirm that the residual error of our algorithm with a random start converges to zero.
For the purposes of benchmarking, we run a fixed number of iterations of the BCD algorithm rather than using a convergence check.

%For the synthetic datasets, our open source code supports (a) low-rank tensor (b) uniform random and (c) positive shifted normal distribution of $\mathfrak{N}(3,1)$ -- that is change the mean such that all the random numbers are positive. We considered three different synthetic matrices for different cases. For baseline comparison with Liavas \cite{LK+17b}, we considered a three mode 
%uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
%mode synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids 
%$1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$ for strong scaling 
%experiments.  In the case of weak scaling of four mode synthetic tensors with (D) and without (N) the use of dimension trees.  The 
%tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$. In all the cases the dimensions were considered such that synthetic tensors can be accommodated even on single node with 64GB for scale up plots.

\subsection{Machine Details}
The entire experimentation was performed on Eos, a supercomputer at the Oak Ridge Leadership Computing Facility. 
Eos is a 736-node Cray XC30 cluster of Intel Xeon E5-2670 processors with a total of 47.104TB of memory. 
Its compute nodes are organized in blades where each blade contains 4 nodes, and every node has 2 sockets with 8 physical cores and 64GB memory. 
The machine support Intel's hyper-threading (HT), but we restricted it because HT offers minimal improvement for BLAS and LAPACK operations. 
In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).

Our objective of the implementation is using open source software as much as possible 
to promote reproducibility and reuse of our code.
%Unlike Liavas \cite{LK+17b} that uses Eigen matrix library \cite{eigenweb}, 
We use Armadillo \cite{sanderson2010} for matrix representation
and operations.  
In Armadillo, the elements of the dense matrix are stored in column major order.
For dense BLAS and LAPACK operations, we linked Armadillo with the default LAPACK/BLAS wrappers from Cray. 
For compiler, we use GNU C++ Compiler (g++ (GCC) 6.3.0) and MPI library is from Cray.  We could also 
compile and run the code in Rhea the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.

\subsection{Comparison Implementations}
The implementation proposed by Liavas et al. \cite{LK+17b} is the only publicly available distributed-memory software (of which we are aware) for computing the CP decomposition of dense tensors, with or without constraints.
We use the acronym NbAO-NTF for Nesterov-based Alternating Optimization Nonnegative Tensor Factorization to refer to their code.

It is based on the same parallel algorithm as our implementation, though it is limited to 3D tensors.
The code uses MPI collectives for communication and Eigen \cite{eigenweb} as an interface to BLAS and LAPACK.
We compiled the code linked to BLAS/LAPACK wrappers from Cray  (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
For fair comparison, we use a flat MPI configuration (one MPI process per core) on all comparisons between the two implementations.

We also point out a difference between the Nesterov-based algorithm and the BPP algorithm for solving the NLS subproblems.
The Nesterov-based algorithm attempts an acceleration step using a linear combination of the current and proposed future step; however, it re-computes the residual error before deciding whether or not to accept or reject the acceleration step.
This residual error cannot always be computed cheaply, using the technique described in \cref{sec:error}, and it contributes significantly (approximately 25\%) to the overall run time.
Because the BPP algorithm does not require this extra computation, and studying convergence behavior of the different NLS algorithms is beyond the scope of this work, we remove the time spent in the acceleration step of NbAO-NTF in all our comparisons.

Our proposed algorithm uses dimension trees, but we also benchmark our implementation without that optimization to highlight its importance.
We use an existing implementation to perform the individual MTTKRPs \cite{HBJT18} with this approach.

\subsection{Strong Scaling}

We perform two strong scaling experiments to compare performance with NbAO-NTF.
The experiments use a cubical synthetic tensor and the HSI image used in \cite{LK+17b}, both of which are 3D.

The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
There is little difference in performance between our implementation without dimension trees and NbAO-NTF.

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_3D_syn.dat}
\renewcommand{\numiterations}{1}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
\label{fig:strongsynthetic3D}
\end{figure}

\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.

%\textcolor{red}{Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$. Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.}

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_3D_HSI.dat}
\renewcommand{\numiterations}{10}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $k \times k\times 1$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
\label{fig:stronghsi3D}
\end{figure}

In \Cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
Because the tensor is 5D, we can no longer compare against NbAO-NTF.
We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from more efficient DGEMM performance and avoiding memory-bound KRP computation.  
That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
Also, for tensors with balanced dimensions, the dimension tree approach yields more favorable shapes for DGEMM.

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_5D_syn.dat}
\renewcommand{\numiterations}{10}
\liavasfalse
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.  The rank is fixed at 32.}
\label{fig:strongsynthetic5D}
\end{figure}

\subsection{Weak Scaling Time Breakdown}

We also perform a weak scaling experiment to understand the time it takes to solve bigger problems with more processors.
In this experiment, we use a synthetic 4D tensor and keep the amount of tensor data assigned to each processor constant, with tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$ and the rank fixed at 32. 
%For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384. 
The results of the breakdown plot is shown in \Cref{fig:weaksynthetic4D}. 
In this case, the algorithm is compute bound with and without the use of the dimension tree, so the total time of the weak scaling remains fixed for both cases. 
Using a dimension tree, the time is completely dominated by the MTTKRP computation.  
Without using a dimension tree, we observe that the KRP is expensive and yields a $2.5\times$ slower total run time even in the 4D case.

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/wk_4D_syn.dat}
\renewcommand{\numiterations}{10}
\weakthreeDfalse
\breakdownplot
%\labels
\end{tikzpicture}
\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.  The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.  The reported times are per iteration.}
\label{fig:weaksynthetic4D}
\end{figure}

\subsection{Varying Processor Grid}

In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.

\begin{figure}
\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}
\begin{axis}[	
	ybar stacked,
	bar width=8pt,
	width=\columnwidth,
	height =.75\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Processor Grid},
	y label style={yshift=-.5cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
	%xtick=data,
	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
\label{fig:commsweep10}
\end{figure}

\subsection{Varying Approximation Rank}

One of the challenges of the CP (and NNCP) decomposition in practice is the choice of decomposition rank.
The most common technique is to compute multiple CP decompositions for various ranks.
As the rank $R$ increases, the approximation error  $\|\TA - \T{M}\|$ decreases with the better approximation power of more parameters. 
However, the benefit of increasing $R$ eventually diminishes if the data can be well approximated with a CP model.
%that an increase in $R$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
%values, like 100 to 105. 
%Hence, it is common practice in the community to sweep $k$, to obtain better approximation error within the 
%manageable computation. 
Towards this end, we experiment with various values of $R$ to observe the relative increase in running time for two real-world data sets.

\Cref{fig:ksweep4DHSI} shows the time breakdown of our implementation using a dimension tree on the 4D HSI dataset for $R=\{10,\dots,50\}$. 
We observe an overall time increase with increased $R$, but each part of the computation scales slightly differently.
The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $R$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).
As $R$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
The local NLS time is increasing with $O(R^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(R^2)$, becoming a significant cost for larger $R$.

In \Cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
Starting at $R=10$ we see the largest speed up of $2\times$ for our implementation with a dimension tree over NbAO-NTF. 
This is due to a combination of the dimension tree performing fewer flops in the MTTKRPs and KRPs. 
However, as the rank increases this speed up diminishes to $1.6\times$. 
The loss of speed up is a result of the fact that, as we observed in \Cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
Again, the performance of our implementation without the dimension tree optimization is comparable to NbAO-NTF.
